---
title: "detectRUNS: an R package to detect runs of homozygosity and heterozygosity in diploid genomes"
author: "Filippo Biscarini, Paolo Cozzi, Gabrtiele Marras"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{detectRUNS: an R package to detect runs of homozygosity and heterozygosity in diploid genomes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- 
PLEASE NOTE

This is a vignette and it's not rebuilt automatically with the package.
To do so please use the following command:

devtools::build_vignettes()

And *then* rebuild the package, so it updates the doc.
--> 

```{r, eval=TRUE, include=TRUE, echo=FALSE}
#just to be sure
library("detectRUNS")
```

## Overview

**detectRUNS** is an R package for the detection of runs of homozygosity (ROH/ROHom) and of heterozygosity (ROHet, a.k.a. "heterozygosity-rich regions") in diploid genomes. detectRUNS uses two methods to detect genomic runs: 

1. sliding-window based method: 
2. consecutive runs:

the performances of one or more genomic regressors when artificial noise is injected in an existing dataset. It is possible, however, to use GROAN to simply compare the prediction accuracy of different regression models without the injection of extra noise. Tests can either be perfomed on a single dataset (through crossvalidation) or multiple datasets (one for training, and one or more for testing). Crossvalidation and data management are transparent and the user is only required high level input.

- Never uses retina figures
- Has a smaller default figure size
- Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style

## Sample data

```{r}
genotypeFilePath <- system.file("extdata", "Kijas2016_Sheep_subset.ped", package="detectRUNS")
mapFilePath <- system.file("extdata", "Kijas2016_Sheep_subset.map", package="detectRUNS")
```

Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
