---
title: "detectRUNS: an R package to detect runs of homozygosity and heterozygosity in diploid genomes"
author: "Filippo Biscarini, Paolo Cozzi, Giustino Gaspa, Gabriele Marras"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{detectRUNS: an R package to detect runs of homozygosity and heterozygosity in diploid genomes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- 
PLEASE NOTE

This is a vignette and it's not rebuilt automatically with the package.
To do so please use the following command:

devtools::build_vignettes()

And *then* rebuild the package, so it updates the doc.
--> 

```{r, eval=TRUE, include=TRUE, echo=FALSE}
#just to be sure
library("detectRUNS")
```

## Overview

**detectRUNS** is a R package for the detection of **runs of homozygosity** (**ROH/ROHom**) and of **heterozygosity** (**ROHet**, a.k.a. "**heterozygosity-rich regions**") in diploid genomes.
**ROH/ROHom** were first studied in humans (e.g. McQuillan et al. 2008) and rapidly found applications not only in human genetics abut also in animal genetics (e.g. Ferencakovic et al., 2011, in *Bos taurus*). More recently, the idea of looking also at "runs of heterozygosity" (**ROHet** or, more appropriately, "heterozygosity-rich regions") has been proposed (Wiliams et al. 2016).

detectRUNS uses two methods to detect genomic runs: 

1. sliding-window based method: 
2. consecutive runs:

The sliding-window based method is similar to what is implemented in the computer package *Plink* (Purcell et al., 2007) [see Bjelland et al., 2013 for a description]. In brief, a sliding window is used to scan the genome, and the characteristics of consecutive windows are used to determine whether a SNP is or not in a run (either ROH/ROHom or ROHet). Parameters for both the sliding window and the run need to be specified.

The "consecutive runs" method is window-free and directly scans the genome SNP by SNP. It was first proposed by Marras et al. (2015). Here, only parameters for the runs need to be specified.

Besides detecting genomic runs (again, either homozygosity or heterozygosity, either sliding-window based or consecutive), and saving results to a data frame of individual runs, **detectRUNS** can:

- plot runs along the genome:
    * plot runs per individual
    * plot stacked (piled) runs
    * plot the % of times each SNP is in a run in the population (per chromosome)
    * Manhattan-like plot of the % of times each SNP is in a run
- plot mean or total run length vs number of runs per individual
- generate summary descriptive statistics on detected runs
- calculate inbreeding based on ROH (*$F_{ROH}$*), genome-wide and chromosome-wide
- plot $F_{ROH}$ per chromosome

The input files for **detectRUNS** are *Plink* **ped/map** files. If one wishes to use this R package only for plots and summary statistics, output files from *Plink* (.hom files) can be easily read into **detectRUNS** through a specific function.

**detectRUNS** can be used with genotype data from any diploid organisms: humans, animals or plants.

## Sample data

To illustrate the functionalities of **detectRUNS**, we use data on sheep (*Ovis aries*) SNP genotypes from the work by Kijas et al. (2016), available on-line through "Dryad" (https://goo.gl/sfAy8k). 
A subset with two breeds ("Jacobs" and "Navajo-Churro", 100 animals) and two chromosomes ($4\,841$ SNPs from OAA 2 and 24) was used.

```{r}
genotypeFilePath <- system.file("extdata", "Kijas2016_Sheep_subset.ped", package="detectRUNS")
mapFilePath <- system.file("extdata", "Kijas2016_Sheep_subset.map", package="detectRUNS")
```

## Detect runs

For the detection of genomic runs, **detectRUNS** uses two main functions:

1. **slidingRUNS.run**: for sliding-window-based detection
2. **consecutiveRUNS.run**: for consecutive-SNP-based detection

Input files are to be passed as paths to files (e.g. /home/Documents/experiment/file.ped/map).

### sliding-window-based run detection

The function *slidingRUNS.run()* accepts in input several parameters: besides the paths (or names) of ped/map files, there are parameters related to the sliding window and parameters related to the genomic runs.

Sliding-window parameters are: *windowSize*, *threshold* (to call a SNP "in run"), *minSNP* (minimum number of homozygous/heterozygous SNP in the window), *maxOppWindow* (maximum number of SNP with opposite genotype: heterozygous/homozygous) and *maxMissWindow* (maximum number of missing genotypes).

Run-related parameters are: *maxGap* (maximum gap between consecutive SNPs, in basepairs -bps), *minLenghtBps* (minimum length of the run, in bps), *minDensity* (number of SNPs every $x$ kilo-basepairs -kbps), *maxOppRun* (maximum number of opposite genotypes in the run), *maxMissRun* (maximum number of missing genotypes in the run).

*ROHet* controls whether runs of homozygosity (ROH/ROHom) or of heterozygosity (heterozygosity-rich regions, ROHet) will be detected. It defaults to **FALSE** (ROH/ROHom).

```{r,results='hide',message=FALSE, cache=FALSE, warning=FALSE}
slidingRuns <- slidingRUNS.run(genotypeFile = genotypeFilePath, 
                               mapFile = mapFilePath, 
                               windowSize = 15, 
                               threshold = 0.05,
                               minSNP = 20, 
                               ROHet = FALSE, 
                               maxOppWindow = 1, 
                               maxMissWindow = 1,
                               maxGap = 10^6, 
                               minLengthBps = 250000, 
                               minDensity = 1/10^3, # SNP/kbps
                               maxOppRun = NULL,
                               maxMissRun = NULL
) 
```

### consecutive SNP-based run detection

The function *consecutiveRUNS.run()* has a similar structure, obviously without the sliding-window parameters.  

```{r,results='hide',message=FALSE, cache=FALSE, warning=FALSE}
consecutiveRuns <- consecutiveRUNS.run(
                                      genotypeFile =genotypeFilePath,
                                      mapFile = mapFilePath,
                                      minSNP = 20,
                                      ROHet = FALSE,
                                      maxGap = 10^6,
                                      minLengthBps = 250000,
                                      maxOppRun = 1,
                                      maxMissRun = 1
)
```

*slidingRUNS.run()* detected **`r nrow(slidingRuns)`** ROH; *consecutiveRUNS.run()* detected **`r nrow(consecutiveRuns)`** ROH.

#### "Runs of heterozygosity" (a.k.a. heterozygosity-rich regions)

By setting **ROHet=TRUE**, runs of heterozygosity (a.k.a. heterozygosity-rich genomic regions) are detected instead. Again, the usert can choose whether to use the sliding-window or the consecutive method.

```{r,results='hide',message=FALSE, cache=FALSE, warning=FALSE}
slidingRuns_het <- slidingRUNS.run(genotypeFile = genotypeFilePath, 
                               mapFile = mapFilePath, 
                               windowSize = 10, 
                               threshold = 0.05,
                               minSNP = 10, 
                               ROHet = TRUE, 
                               maxOppWindow = 2, 
                               maxMissWindow = 1,
                               maxGap = 10^6, 
                               minLengthBps = 10000, 
                               minDensity = 1/10^6, # SNP/kbps
                               maxOppRun = NULL,
                               maxMissRun = NULL
) 
```

```{r,results='hide',message=FALSE, cache=FALSE}
consecutiveRuns_het <- consecutiveRUNS.run(
                                      genotypeFile =genotypeFilePath,
                                      mapFile = mapFilePath,
                                      minSNP = 10,
                                      ROHet = TRUE,
                                      maxGap = 10^6,
                                      minLengthBps = 10000,
                                      maxOppRun = 2,
                                      maxMissRun = 1
)
```
*slidingRUNS.run()* detected **`r nrow(slidingRuns_het)`** ROH; *consecutiveRUNS.run()* detected **`r nrow(consecutiveRuns_het)`** ROHet.

Runs of homozygosity (ROH) detected using the sliding-windows method (output from *slidingRUNS.run()*) will be used to illustrate summary statistics, plots and inbreeding calculations.

## Summary statistics on detected runs

The function *summaryRuns()* takes in input the dataframe with results from runs detection and calculates a number of basic descriptive statistics on runs. Additional necessary parameters are the paths to the *Plink* ped and map files. `Class` and `snpInRuns` are optional arguments.

```{r, results='hide',message=FALSE, cache=FALSE}
summaryList <- summaryRuns(runs = slidingRuns, mapFile = mapFilePath, genotypeFile = genotypeFilePath, Class = 6, snpInRuns = TRUE)
```
The returned list includes the following dataframes: 

`r names(summaryList)`

We can, for instance, have a look at the number of runs per class-size (Mbps) in the two breeds: we see that in Jacobs sheep there are `r summaryList$summary_ROH_count[1,1]` ROH with size up to 6 Mbps.

```{r}
summaryList$summary_ROH_count
```

Or, the average number of ROH per chromosome and per breed can be obtained.

```{r}
summaryList$summary_ROH_mean_chr
```

The dataframe "SNPinRun" contains, for each SNP, the proportion of times it falls inside a run in any given population/group:

```{r}
head(summaryList$SNPinRun)
```

The summary information included in the list returned from *summaryRuns()* is conveniently organized in data.frames, so that it can easily be visualized, manipulated and written out to text files (e.g. .csv files).

## Plots

**detectRUNS** produces a number of plots from the dataframe with runs (results from sliding-window or consecutive scans of the genome for ROH/ROHet).
The basic plot, produced by the function *plot_Runs()*

```{r, fig.show='hold', fig.cap="Plot runs per individual"}
slidingRuns[slidingRuns$group=="Navajo-Churro","group"] <- "NC" #rename Navajo_Churro to NC (for plotting convenience)
plot_Runs(runs = slidingRuns)
```

```{r, fig.show='hold', fig.cap="Plot runs per individual"}
plot_StackedRuns(runs = slidingRuns)
```

```{r, fig.show='hold', fig.cap="Plot runs per individual"}
plot_SnpsInRuns(runs = slidingRuns, genotypeFile = genotypeFilePath, mapFile = mapFilePath)
```

## $F_{ROH}$: ROH-based inbreeding

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))

## Importing data from *Plink*

## References

- Bjelland, D. W., K. A. Weigel, N. Vukasinovic, and J. D. Nkrumah. "Evaluation of inbreeding depression in Holstein cattle using whole-genome SNP markers and alternative measures of genomic inbreeding." journal of dairy Science 96, no. 7 (2013): 4697-4706.
- Ferencakovic, Maja, Edin Hamzic, Birgit Gredler, Ino Curik, and Johann Sölkner. "Runs of homozygosity reveal genome-wide autozygosity in the Austrian Fleckvieh cattle." Agriculturae Conspectus Scientificus (ACS) 76, no. 4 (2011): 325-329.
- Kijas, James W., Tracy Hadfield, Marina Naval Sanchez, and Noelle Cockett. "Genome‐wide association reveals the locus responsible for four‐horned ruminant." Animal genetics 47, no. 2 (2016): 258-262.
- Marras, Gabriele, Giustino Gaspa, Silvia Sorbolini, Corrado Dimauro, Paolo Ajmone‐Marsan, Alessio Valentini, John L. Williams, and Nicolò PP Macciotta. "Analysis of runs of homozygosity and their relationship with inbreeding in five cattle breeds farmed in Italy." Animal genetics 46, no. 2 (2015): 110-121.
- McQuillan, Ruth, Anne-Louise Leutenegger, Rehab Abdel-Rahman, Christopher S. Franklin, Marijana Pericic, Lovorka Barac-Lauc, Nina Smolej-Narancic et al. "Runs of homozygosity in European populations." The American Journal of Human Genetics 83, no. 3 (2008): 359-372.
- Purcell, Shaun, Benjamin Neale, Kathe Todd-Brown, Lori Thomas, Manuel AR Ferreira, David Bender, Julian Maller et al. "PLINK: a tool set for whole-genome association and population-based linkage analyses." The American Journal of Human Genetics 81, no. 3 (2007): 559-575.
- Williams, John L., Stephen JG Hall, Marcello Del Corvo, K. T. Ballingall, L. I. C. I. A. Colli, P. A. O. L. O. Ajmone Marsan, and F. Biscarini. "Inbreeding and purging at the genomic Level: the Chillingham cattle reveal extensive, non‐random SNP heterozygosity." Animal genetics 47, no. 1 (2016): 19-27.
