---
title: "detectRUNS: an R package to detect runs of homozygosity and heterozygosity in diploid genomes"
author:
  - name: "Filippo Biscarini, Paolo Cozzi, Giustino Gaspa, Gabriele Marras"
    affiliation: "IBBA-CNR, PTP, Università degli Studi di Sassari, University of Guelph"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{detectRUNS: detecting runs of homozygosity and heterozygosity in diploid genomes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  message   = FALSE,
  warning   = FALSE,
  fig.width = 6,
  fig.height = 4
)
library(detectRUNS)
```

# Overview

**detectRUNS** is an R package for the detection of **Runs of Homozygosity** (**ROH/ROHom**) and of **Runs of Heterozygosity** (**ROHet**, a.k.a. "heterozygosity-rich regions") in diploid genomes.

ROH/ROHom were first studied in humans (McQuillan et al. 2008) and rapidly found applications in animal genetics (e.g. Ferencakovic et al. 2011 in *Bos taurus*). More recently, the idea of looking at "runs of heterozygosity" (ROHet, or more precisely "heterozygosity-rich regions") was proposed (Williams et al. 2016).

detectRUNS supports two detection methods:

1. **Sliding-window method** — similar to the approach implemented in PLINK `--homozyg` (Bjelland et al. 2013). A sliding window scans the genome and the characteristics of consecutive overlapping windows determine whether a SNP is in a run.
2. **Consecutive method** — window-free, SNP-by-SNP scan first proposed by Marras et al. (2015). Only run-level parameters need to be specified.

Both methods are implemented in a high-performance C++ engine that reads PLINK binary (BED/BIM/FAM) files natively and runs in parallel using OpenMP. PLINK text (PED/MAP) files are also supported.

Beyond detecting genomic runs, **detectRUNS** can:

- Plot runs along the genome (per individual, stacked, SNP-in-run frequency)
- Generate descriptive statistics on detected runs
- Calculate ROH-based inbreeding coefficients ($F_{ROH}$), genome-wide and chromosome-wide
- Plot $F_{ROH}$ per chromosome
- Import run results from PLINK, BCFtools, or a previous detectRUNS session

**detectRUNS** can be used with genotype data from any diploid organism: humans, animals or plants.

# Sample data

To illustrate the functionalities of **detectRUNS**, we use data on sheep (*Ovis aries*) SNP genotypes from Kijas et al. (2016). A subset with two breeds ("Jacobs" and "Navajo-Churro", 100 animals) and two chromosomes (4,841 SNPs from OAR 2 and 24) is included in the package.

```{r file-paths}
bedFile <- system.file("extdata", "Kijas2016_Sheep_subset.bed",
                        package = "detectRUNS")
pedFile <- system.file("extdata", "Kijas2016_Sheep_subset.ped",
                        package = "detectRUNS")
mapFile <- system.file("extdata", "Kijas2016_Sheep_subset.map",
                        package = "detectRUNS")
```

# Detecting runs with `scanRUNS()`

The main entry point is `scanRUNS()`. It auto-detects the file format from the extension (`.bed` → BED engine, `.ped` → PED engine) and returns an `ROH` S3 object containing the detected runs and all metadata for downstream analysis.

## Sliding-window method (BED input)

The sliding-window method scans the genome with an overlapping window of `windowSize` SNPs. For each SNP, the fraction of windows in which the SNP falls in a run-compatible state (homozygous for ROHom, heterozygous for ROHet) is computed. If this fraction exceeds `1 - threshold`, the SNP is called "in run". Consecutive SNPs in run are merged into a single run if they satisfy the run-level filters.

Key parameters:

| Parameter | Description | Default |
|---|---|---|
| `windowSize` | Number of SNPs in the sliding window | 15 |
| `threshold` | Max fraction of non-run windows to still call a SNP in run | 0.05 |
| `minSNP` | Minimum number of SNPs in a run | 3 |
| `maxOpp` | Max opposite-genotype SNPs allowed in the window | 1 |
| `maxMiss` | Max missing genotypes allowed in the window | 1 |
| `maxGap` | Max gap (bp) between consecutive SNPs in a run | 10^6 |
| `minLengthBps` | Minimum run length (bp) | 1000 |
| `ROHet` | Detect runs of heterozygosity instead of homozygosity | FALSE |

```{r scan-sliding, results='hide'}
roh_sliding <- scanRUNS(
  bedFile,
  method       = "sliding",
  minSNP       = 20,
  maxOpp       = 1,
  maxMiss      = 1,
  minLengthBps = 250000,
  maxGap       = 1e6,
  windowSize   = 15,
  threshold    = 0.05,
  verbose      = FALSE
)
```

```{r print-sliding}
print(roh_sliding)
```

## Consecutive method (PED input)

The consecutive method scans SNPs one by one, without a sliding window. A run starts when a SNP satisfies the genotype criterion (homozygous for ROHom) and ends when the criterion is violated. Run-level filters then apply.

Key parameters (no window parameters needed):

| Parameter | Description | Default |
|---|---|---|
| `minSNP` | Minimum number of SNPs in a run | 3 |
| `maxOpp` | Max opposite-genotype SNPs allowed in the run | 0 |
| `maxMiss` | Max missing genotypes allowed in the run | 0 |
| `maxGap` | Max gap (bp) between consecutive SNPs | 10^6 |
| `minLengthBps` | Minimum run length (bp) | 1000 |

```{r scan-consecutive, results='hide'}
roh_cons <- scanRUNS(
  pedFile,
  method       = "consecutive",
  minSNP       = 20,
  maxOpp       = 1,
  maxMiss      = 1,
  minLengthBps = 250000,
  maxGap       = 1e6,
  verbose      = FALSE
)
```

```{r print-consecutive}
print(roh_cons)
```

The sliding-window method detected **`r nrow(roh_sliding$runs)`** ROH; the consecutive method detected **`r nrow(roh_cons$runs)`** ROH.

## Runs of heterozygosity (ROHet)

By setting `ROHet = TRUE`, heterozygosity-rich regions are detected instead of homozygosity runs. These are genomic regions where an individual is heterozygous at an unusually high proportion of loci — potentially indicating balancing selection, recent admixture, or structural variation.

```{r scan-rohet, results='hide'}
roh_het <- scanRUNS(
  bedFile,
  method       = "sliding",
  ROHet        = TRUE,
  minSNP       = 10,
  maxOpp       = 2,
  maxMiss      = 1,
  minLengthBps = 10000,
  maxGap       = 1e6,
  windowSize   = 10,
  threshold    = 0.05,
  verbose      = FALSE
)
```

`r nrow(roh_het$runs)` ROHet regions were detected across the two chromosomes.

# The ROH object

`scanRUNS()` returns an `ROH` S3 object — a named list carrying the detected runs and all associated metadata. Once you have an `ROH` object, no file paths are needed for any downstream function.

```{r roh-structure}
names(roh_sliding)
```

| Element | Content |
|---|---|
| `$runs` | One row per run: group, id, chrom, nSNP, from, to, lengthBps |
| `$summary` | Per-individual summary: n_ROH, total_length_bp, mean_length |
| `$chrom_lengths` | Chromosome lengths derived from the SNP map |
| `$sample_info` | All samples (group + id), including those with 0 runs |
| `$snp_map` | Full SNP map (CHR, SNP_NAME, POSITION) |
| `$method` | Detection method |
| `$type` | ROHom or ROHet |
| `$snp_freq` | Per-SNP ROH frequency (BED engine only) |

```{r roh-access}
head(as.data.frame(roh_sliding), 4)
```

```{r sample-info}
head(roh_sliding$sample_info)
table(roh_sliding$sample_info$group)
```

## Save and reload results

Large datasets can take minutes to scan. Save the result once and reload instantly in future sessions:

```{r save-load, eval=FALSE}
saveROH(roh_sliding, "sheep_roh.roh")
roh_sliding <- loadROH("sheep_roh.roh")
```

# Summary statistics

`summaryRuns()` computes descriptive statistics on detected runs. It accepts an `ROH` object directly — no file paths needed.

```{r summary, results='hide'}
summ <- summaryRuns(roh_sliding, Class = 6, snpInRuns = TRUE)
```

The returned list includes `r length(summ)` data frames:

```{r summary-names}
names(summ)
```

For instance, the number of runs per size-class (Mbps) per breed:

```{r summary-count}
summ$summary_ROH_count
```

The average ROH length (Mb) per chromosome and breed:

```{r summary-mean-chr}
summ$summary_ROH_mean_chr
```

The `SNPinRun` element gives, for each SNP and each population, the proportion of individuals that carry a run overlapping that SNP:

```{r summary-snpinrun}
head(summ$SNPinRun)
```

`tableRuns()` identifies genomic regions where a run is shared by a high fraction of individuals — useful for detecting selection signatures:

```{r table-runs}
topRuns <- tableRuns(roh_sliding, threshold = 0.7)
print(topRuns)
```

# Inbreeding ($F_{ROH}$)

Individual inbreeding coefficients from ROH are:

$$F_{ROH} = \frac{\sum L_{ROH}}{L_{genome}}$$

where $\sum L_{ROH}$ is the total length of all ROH in an individual and $L_{genome}$ is the total surveyed genome length.

```{r froh}
froh <- Froh_inbreeding(roh_sliding, genome_wide = TRUE)
head(froh[order(froh$Froh_genome, decreasing = TRUE), ], 6)
```

Setting `genome_wide = FALSE` returns per-chromosome estimates:

```{r froh-chr}
froh_chr <- Froh_inbreeding(roh_sliding, genome_wide = FALSE)
head(froh_chr, 4)
```

Inbreeding can also be partitioned by ROH length class, which reflects the age of the inbreeding event (short ROH = ancient, long ROH = recent):

```{r froh-class}
froh_class <- Froh_inbreedingClass(roh_sliding, Class = 6)
head(froh_class, 4)
```

# Plots

All plot functions accept the `ROH` object directly.

## Runs per individual

`plot_Runs()` plots detected runs for each individual along the chromosome. Different groups/populations are coloured differently, making it easy to contrast patterns between breeds.

```{r plot-runs, fig.cap="Detected ROH per individual along the chromosome."}
plot_Runs(roh_sliding, suppressInds = FALSE, savePlots = FALSE)
```

## Stacked runs

`plot_StackedRuns()` stacks all runs on top of each other, making regions with an excess of ROH visually prominent. Separate plots are produced per chromosome and per group.

```{r plot-stacked, fig.cap="Stacked ROH — regions enriched for runs stand out clearly.", results='hide'}
plot_StackedRuns(roh_sliding, savePlots = FALSE)
```

## SNP-in-run frequency

`plot_SnpsInRuns()` plots, for each SNP, the proportion of individuals in each group that carry a run overlapping that SNP position. This highlights genomic regions under selection or with population-specific autozygosity.

```{r plot-snps-chr2, fig.cap="Proportion of individuals with a run at each SNP — OAR2.", results='hide'}
plot_SnpsInRuns(
  runs    = roh_sliding$runs[roh_sliding$runs$chrom == 2, ],
  mapFile = mapFile,
  genotypeFile = pedFile,
  savePlots = FALSE
)
```

```{r plot-snps-chr24, fig.cap="Proportion of individuals with a run at each SNP — OAR24.", results='hide'}
plot_SnpsInRuns(
  runs    = roh_sliding$runs[roh_sliding$runs$chrom == 24, ],
  mapFile = mapFile,
  genotypeFile = pedFile,
  savePlots = FALSE
)
```

In the Jacob breed a clear peak of ROH is visible approximately halfway along OAR2. This corresponds to the strong GWAS signal reported by Kijas et al. (2016) on OAR2 associated with the four-horns phenotype — a striking example of selection leaving a footprint of autozygosity.

## Manhattan plot of SNP-in-run frequency

`plot_manhattanRuns()` displays the SNP-in-run frequency across all chromosomes simultaneously, in the style of a GWAS Manhattan plot:

```{r plot-manhattan, fig.cap="Manhattan plot of SNP-in-run frequency (Jacobs breed).", results='hide'}
plot_manhattanRuns(
  runs         = roh_sliding$runs[roh_sliding$runs$group == "Jacobs", ],
  genotypeFile = pedFile,
  mapFile      = mapFile,
  savePlots    = FALSE
)
```

## Run length distribution

`plot_DistributionRuns()` visualises the distribution of run lengths in several styles (`MeanClass`, `MeanChr`, `RunsPCT`, `RunsPCT_Chr`, or `All`):

```{r plot-distribution, fig.cap="Run length distribution by group and class.", results='hide'}
plot_DistributionRuns(roh_sliding, style = "MeanClass", Class = 6,
                      savePlots = FALSE)
```

## Inbreeding per chromosome

`plot_InbreedingChr()` plots $F_{ROH}$ per chromosome, for example as a box-plot by group:

```{r plot-inbreeding-chr, fig.cap="Froh per chromosome by breed (box-plot)."}
plot_InbreedingChr(roh_sliding, style = "FrohBoxPlot", savePlots = FALSE)
```

## Violin plot of run length vs number of runs

`plot_ViolinRuns()` shows the distribution of total or mean run length against the number of runs per individual, useful for comparing inbreeding patterns between populations:

```{r plot-violin, fig.cap="Violin plot: total ROH length by group."}
plot_ViolinRuns(roh_sliding, method = "sum", savePlots = FALSE)
```

# Importing external results

Run results from PLINK (`--homozyg`), BCFtools (`roh`), or a previous detectRUNS session saved to CSV can be imported with `readExternalRuns()` and wrapped into an `ROH` object using `as_ROH()`:

```{r import}
runsFile <- system.file("extdata", "Kijas2016_Sheep_subset.sliding.csv",
                         package = "detectRUNS")
ext_runs <- readExternalRuns(inputFile = runsFile, program = "detectRUNS")
head(ext_runs, 3)
```

```{r as-roh}
roh_ext <- as_ROH(ext_runs,
                  mapFile      = mapFile,
                  genotypeFile = pedFile,
                  method       = "sliding",
                  type         = "ROHom")
print(roh_ext)
```

# Backward compatibility

Code written for earlier versions of detectRUNS continues to work. The deprecated functions `slidingRUNS.run()` and `consecutiveRUNS.run()` still exist and return plain data.frames. All statistics and plot functions still accept a plain data.frame alongside explicit `mapFile=` and `genotypeFile=` arguments:

```{r backward-compat, eval=FALSE}
# Old API — still works, but deprecated
old_runs <- suppressWarnings(
  slidingRUNS.run(pedFile, mapFile,
                  windowSize = 15, threshold = 0.05,
                  minSNP = 20, minLengthBps = 250000)
)
Froh_inbreeding(old_runs, mapFile = mapFile)
plot_Runs(old_runs)
```

# References

- Bjelland, D.W., Weigel, K.A., Vukasinovic, N., Nkrumah, J.D. (2013). Evaluation of inbreeding depression in Holstein cattle using whole-genome SNP markers and alternative measures of genomic inbreeding. *Journal of Dairy Science*, 96(7):4697–4706.
- Ferencakovic, M., Hamzic, E., Gredler, B., Curik, I., Sölkner, J. (2011). Runs of homozygosity reveal genome-wide autozygosity in the Austrian Fleckvieh cattle. *Agriculturae Conspectus Scientificus*, 76(4):325–329.
- Kijas, J.W., Hadfield, T., Naval Sanchez, M., Cockett, N. (2016). Genome-wide association reveals the locus responsible for four-horned ruminant. *Animal Genetics*, 47(2):258–262.
- Marras, G., Gaspa, G., Sorbolini, S., Dimauro, C., Ajmone-Marsan, P., Valentini, A., Williams, J.L., Macciotta, N.P.P. (2015). Analysis of runs of homozygosity and their relationship with inbreeding in five cattle breeds farmed in Italy. *Animal Genetics*, 46(2):110–121.
- McQuillan, R., Leutenegger, A.L., Abdel-Rahman, R., Franklin, C.S., et al. (2008). Runs of homozygosity in European populations. *The American Journal of Human Genetics*, 83(3):359–372.
- Purcell, S., Neale, B., Todd-Brown, K., Thomas, L., et al. (2007). PLINK: a tool set for whole-genome association and population-based linkage analyses. *The American Journal of Human Genetics*, 81(3):559–575.
- Williams, J.L., Hall, S.J.G., Del Corvo, M., Ballingall, K.T., Colli, L., Ajmone Marsan, P., Biscarini, F. (2016). Inbreeding and purging at the genomic level: the Chillingham cattle reveal extensive, non-random SNP heterozygosity. *Animal Genetics*, 47(1):19–27.
