---
title: "detectRUNS: detecting runs of homozygosity and heterozygosity in diploid genomes"
author: "Filippo Biscarini, Paolo Cozzi, Giustino Gaspa, Gabriele Marras"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{detectRUNS: detecting runs of homozygosity and heterozygosity in diploid genomes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  message   = FALSE,
  warning   = FALSE
)
library(detectRUNS)
```

# Overview

**detectRUNS** is an R package for detecting **Runs of Homozygosity** (ROH/ROHom) and **Runs of Heterozygosity** (ROHet) in diploid genomes. It supports two detection methods:

1. **Sliding-window method** — similar to PLINK `--homozyg` (Bjelland et al. 2013)
2. **Consecutive method** — window-free, SNP-by-SNP scan (Marras et al. 2015)

Both methods are implemented in a high-performance C++ engine that reads PLINK binary (BED/BIM/FAM) files directly and runs in parallel using OpenMP. PLINK text (PED/MAP) files are also supported.

# Quick start

The main entry point is `scanRUNS()`. It auto-detects the file format from the file extension and returns an `ROH` object containing the detected runs and all metadata needed for downstream analysis.

```{r file-paths}
bedFile <- system.file("extdata", "Kijas2016_Sheep_subset.bed",
                        package = "detectRUNS")
pedFile <- system.file("extdata", "Kijas2016_Sheep_subset.ped",
                        package = "detectRUNS")
mapFile <- system.file("extdata", "Kijas2016_Sheep_subset.map",
                        package = "detectRUNS")
```

## Detect ROH with the sliding-window method (BED input)

```{r scan-sliding, results='hide'}
roh <- scanRUNS(
  bedFile,
  method       = "sliding",
  minSNP       = 15,
  maxOpp       = 1,
  maxMiss      = 1,
  minLengthBps = 100000,
  maxGap       = 1e6,
  windowSize   = 15,
  threshold    = 0.05,
  verbose      = FALSE
)
```

```{r print-roh}
print(roh)
```

## Detect ROH with the consecutive method (PED input)

```{r scan-consecutive, results='hide'}
roh_cons <- scanRUNS(
  pedFile,
  method       = "consecutive",
  minSNP       = 15,
  maxOpp       = 0,
  maxMiss      = 0,
  minLengthBps = 100000,
  maxGap       = 1e6,
  verbose      = FALSE
)
```

## Detect ROHet (runs of heterozygosity)

Set `ROHet = TRUE` to detect heterozygosity-rich regions instead:

```{r scan-rohet, results='hide'}
roh_het <- scanRUNS(
  bedFile,
  method       = "sliding",
  ROHet        = TRUE,
  minSNP       = 10,
  maxOpp       = 2,
  maxMiss      = 1,
  minLengthBps = 10000,
  maxGap       = 1e6,
  verbose      = FALSE
)
```

# The ROH object

`scanRUNS()` returns an `ROH` S3 object — a named list carrying the detected runs and all associated metadata:

```{r roh-structure}
names(roh)
```

| Element | Content |
|---|---|
| `$runs` | One row per detected run: group, id, chrom, nSNP, from, to, lengthBps |
| `$summary` | Per-individual summary: n_ROH, total_length_bp, mean_length, n_snps_in_roh |
| `$chrom_lengths` | Chromosome lengths derived from the map |
| `$sample_info` | All individuals (group + id), including those with 0 runs |
| `$snp_map` | Full SNP map: CHR, SNP_NAME, POSITION |
| `$method` | Detection method used |
| `$type` | ROHom or ROHet |
| `$snp_freq` | Per-SNP ROH frequency (BED engine only) |

```{r roh-access}
head(as.data.frame(roh), 3)
```

## Save and reload results

Large datasets can take minutes to scan. Save the `ROH` object once and reload it instantly in future sessions:

```{r save-load, eval=FALSE}
saveROH(roh, "mydata_roh.roh")
roh <- loadROH("mydata_roh.roh")
```

# Sample data

The example dataset is a subset of sheep (*Ovis aries*) SNP genotypes from Kijas et al. (2016): 100 animals from two breeds (Jacobs and Navajo-Churro), 4,841 SNPs on chromosomes 2 and 24.

```{r sample-info}
head(roh$sample_info)
table(roh$sample_info$group)
```

# Summary statistics

`summaryRuns()` computes descriptive statistics on detected runs. When an `ROH` object is passed, no file paths are needed:

```{r summary, results='hide'}
summ <- summaryRuns(roh_cons, Class = 2, snpInRuns = TRUE)
```

```{r summary-show}
names(summ)
summ$summary_ROH_count
summ$summary_ROH_mean_chr
head(summ$SNPinRun, 3)
```

The `SNPinRun` element gives, for each SNP, the proportion of times it falls inside a run in each population/group.

`tableRuns()` reports genomic regions where a high proportion of individuals share a run — useful for identifying selection signatures:

```{r table-runs}
topRuns <- tableRuns(roh_cons, threshold = 0.5)
head(topRuns, 5)
```

# Inbreeding ($F_{ROH}$)

Individual inbreeding coefficients based on ROH are calculated as:

$$F_{ROH} = \frac{\sum L_{ROH}}{L_{genome}}$$

```{r froh}
froh <- Froh_inbreeding(roh_cons, genome_wide = TRUE)
head(froh[order(froh$Froh_genome, decreasing = TRUE), ], 5)
```

Chromosome-level inbreeding:

```{r froh-chr}
froh_chr <- Froh_inbreeding(roh_cons, genome_wide = FALSE)
head(froh_chr, 3)
```

Inbreeding can also be summarised by ROH length class:

```{r froh-class}
froh_class <- Froh_inbreedingClass(roh_cons, Class = 2)
head(froh_class, 3)
```

# Plots

All plot functions accept the `ROH` object directly:

```{r plots, eval=FALSE}
# Individual runs per chromosome
plot_Runs(roh_cons, suppressInds = TRUE, savePlots = FALSE)

# Stacked runs (regions with excess ROH are easily visible)
plot_StackedRuns(roh_cons, savePlots = FALSE)

# Run length distribution by group
plot_ViolinRuns(roh_cons, method = "sum", savePlots = FALSE)

# Run length distribution and Froh by class
plot_DistributionRuns(roh_cons, style = "All", Class = 2, savePlots = FALSE)

# Froh per chromosome
plot_InbreedingChr(roh_cons, style = "FrohBoxPlot", savePlots = FALSE)

# SNP-in-run frequency per chromosome
plot_SnpsInRuns(roh_cons, savePlots = FALSE)

# Manhattan-style plot of SNP-in-run frequency
plot_manhattanRuns(roh_cons, savePlots = FALSE)
```

# Importing external results

Run results from PLINK (`--homozyg`), BCFtools (`roh`), or a previous detectRUNS session saved to CSV can be imported via `readExternalRuns()` and wrapped into an `ROH` object with `as_ROH()`:

```{r import}
runsFile <- system.file("extdata", "Kijas2016_Sheep_subset.sliding.csv",
                         package = "detectRUNS")
ext_runs <- readExternalRuns(runsFile, program = "detectRUNS")
head(ext_runs, 3)
```

```{r as-roh}
roh_ext <- as_ROH(ext_runs,
                  mapFile      = mapFile,
                  genotypeFile = pedFile,
                  method       = "sliding",
                  type         = "ROHom")
print(roh_ext)
```

# Backward compatibility

Code written for previous versions of detectRUNS continues to work. The deprecated functions `slidingRUNS.run()` and `consecutiveRUNS.run()` still exist and return plain data.frames. All statistics and plot functions still accept a plain data.frame alongside explicit `mapFile=` and `genotypeFile=` arguments:

```{r backward-compat, eval=FALSE}
# Old-style (still works, but deprecated)
old_runs <- suppressWarnings(
  slidingRUNS.run(pedFile, mapFile, windowSize = 15, threshold = 0.05,
                  minSNP = 15, minLengthBps = 100000)
)
Froh_inbreeding(old_runs, mapFile = mapFile)
```

# References

- Bjelland, D.W. et al. (2013). *J. Dairy Sci.* 96(7):4697–4706.
- Kijas, J.W. et al. (2016). *Animal Genetics* 47(2):258–262.
- Marras, G. et al. (2015). *Animal Genetics* 46(2):110–121.
- McQuillan, R. et al. (2008). *Am. J. Hum. Genet.* 83(3):359–372.
- Purcell, S. et al. (2007). *Am. J. Hum. Genet.* 81(3):559–575.
- Williams, J.L. et al. (2016). *Animal Genetics* 47(1):19–27.
